# Elementary Statistics - Quick Review
<div style='float: right; font-size:10px; margin-top: -20px; color: #ccc;'>Rafe C. H. Liu, LastUpdate: 2018-07-18</div>
  
## Preliminaries

- Necessaries
    - [Quick R](http://www.statmethods.net/index.html)
    - [Elementary Statistics with R](http://www.r-tutor.com/elementary-statistics)
    - [idre - R](https://stats.idre.ucla.edu/r/) 
    
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
options(width = 150)

ubike <- read.csv(file = "./data/ubike-weather-big5.csv",
          colClasses = c("factor","integer","integer","factor","factor",
                         "numeric","numeric","integer","numeric","integer",
                         "integer","numeric","numeric", "integer","integer",
                         "numeric","numeric","numeric", "numeric","numeric",
                         "numeric", "numeric","numeric"), 
          fileEncoding = 'BIG-5',
          stringsAsFactors = F)


ubike <- ubike %>% select(
  date, hour, sno, sarea, sna, lat, lng, tot, 
  sbi = avg.sbi, bemp = avg.bemp,
  temp, humidity, pressure
) %>% mutate(
  sbi = ceiling(sbi), 
  bemp = ceiling(bemp),
  temp = temp %>% round(digits = 2), 
  humidity = humidity %>% round(digits = 2), 
  pressure = pressure %>% round(digits = 2)
)

ubike <- ubike %>% 
  mutate(date = as.Date(date),
         sarea = as.factor(sarea))

###

chin <- c("日期","時間","場站代號","場站區域","場站名稱","經度","緯度",
          "總停車格","車輛數","空位數","氣溫","溼度","氣壓")

coln <- data.frame(matrix(data="", ncol=4, nrow=7), stringsAsFactors = FALSE)
coln$X1 <- chin[1:7]
coln$X2 <- colnames(ubike)[1:7]
coln$X3[1:6] <- chin[8:13]
coln$X4[1:6] <- colnames(ubike)[8:13]
```

- Dataset for this tutorial
    - **aggregate** vs disaggregate data
    - aggregate level

```{r eval=FALSE}
# import dataset from file
ubike <- read.csv(file = "./data/ubike-weather-big5.csv", 
                  fileEncoding = 'BIG-5',
                  stringsAsFactors = F)

# A galance of dataset
head(ubike)
tail(ubike)
str(ubike)
summary(ubike)

# basic Extract-Transform-Load process using dplyr
library(dplyr)

ubike <- ubike %>% select(
  date, hour, sno, sarea, sna, lat, lng, tot, 
  sbi = avg.sbi, bemp = avg.bemp,
  temp, humidity, pressure
) %>% mutate(
  sbi = ceiling(sbi), 
  bemp = ceiling(bemp),
  temp = temp %>% round(digits = 2), 
  humidity = humidity %>% round(digits = 2), 
  pressure = pressure %>% round(digits = 2)
)

# check again ==> correct variable types
# str(ubike)
ubike <- ubike %>% 
  mutate(date = as.Date(date),
         sarea = as.factor(sarea))
```

<br>

```{r echo=FALSE}
knitr::kable(coln, col.names = c("欄位名稱","代號","欄位名稱","代號"))
```


## Descriptive Statistics

```{r fig.height=4, fig.width=12}
summary(ubike)

ubike %>%
  group_by(sarea, sna) %>%
  summarise(mean(bemp), median(bemp), max(bemp), min(bemp), sd(bemp))

# Combine plots
par(mfrow=c(1,3)) # plot parameter mfrow: number of Multiple Figures (use ROW-wise).
hist(ubike$bemp) # hist plot
boxplot(ubike$bemp, horizontal = TRUE) # boxplot
plot(density(ubike$bemp)) # kernel density plots
par(mfrow=c(1,1)) # reset
```

### Cross-Analysis

- Summary Statistics by Groups

```{r}
library(dplyr)

## how many stations by districts 
ubike %>%
  filter(date == '2014-12-08', hour == '20') %>%
  group_by(sarea) %>%
  summarise(stations = n()) %>%
  head()

## how many rented bikes by station per hour
# Long-data format to Wide-data
library(reshape2)

ubike %>%
  filter(sna == '世貿二館' | sna == '信義廣場(台北101)') %>%
  # filter(sna %in% c('世貿二館', '信義廣場(台北101)')) %>%
  select(date, hour, sna, sbi) %>%
  dcast(formula = date + hour ~ sna, value.var = 'sbi') %>%
  head()
```

### Correlation

- References
    - http://www.statmethods.net/stats/correlations.html

```{r}
dat <- ubike %>%
  filter(sna == '世貿二館' | sna == '信義廣場(台北101)') %>%
  # filter(sna %in% c('世貿二館', '信義廣場(台北101)')) %>%
  select(date, hour, sna, sbi) %>%
  dcast(formula = date + hour ~ sna, value.var = 'sbi')

cov(dat[,3:4])
cor(dat[,3:4], method = 'pearson')
```


## Exploratory Data Analysis

```{r}
ubike %>%
  filter(sna == '世貿二館' | sna == '信義廣場(台北101)') %>%
  # filter(sna %in% c('世貿二館', '信義廣場(台北101)')) %>%
  select(date, hour, sarea, sna, sbi) %>%
  head(10)

## plots
library(ggplot2)

ubike %>%
  filter(sna == '世貿二館' | sna == '信義廣場(台北101)') %>%
  select(date, hour, sarea, sna, sbi) %>% 
  ggplot(aes(x = hour, y = sbi, group = sna, color = sna)) +
  geom_line() + facet_grid(date~.) + 
  theme(text = element_text(family = 'STHeiti')) # for MAC user
```

## Hypothesis Testing

- References
    - [Which Statistical Test?](https://stats.idre.ucla.edu/other/mult-pkg/whatstat/)
    - [Quick-R: ANOVA & MANOVA](http://www.statmethods.net/stats/anova.html)
    - [變異數分析 ANOVA (作者：陳鍾誠)](http://programmermagazine.github.io/201311/htm/article3.html)

### t-tests

- Quiz: 檢定兩區域平均租借數是否相等 (Independent 2-group sample t-test)
    - H0: u1 = u2 vs H1: u1 != u2 
    - p is small, reject H0, u1 != u2

```{r}
# Dataset
dat <- ubike %>%
  filter(sna == '世貿二館' | sna == '信義廣場(台北101)') %>%
  # filter(sna %in% c('世貿二館', '信義廣場(台北101)')) %>%
  select(date, hour, sna, sbi) %>%
  dcast(formula = date + hour ~ sna, value.var = 'sbi')

# t-test
t.test(dat$世貿二館, dat$`信義廣場(台北101)`) 

## results are stored in the list
rt <- t.test(dat$世貿二館, dat$`信義廣場(台北101)`) 
rt$p.value
```


### ANOVA

- Quiz: 探討四車站(sna)的平均租借數(sbi)租借數趨勢**

```{r}
# Dataset
dat <- ubike %>%
  filter(sna %in% c('世貿二館', '世貿三館', '信義廣場(台北101)', '市民廣場')) %>%
  select(date, hour, sna, sbi) 

head(dat)
summary(dat)

# EDA
dat %>% 
  ggplot(aes(x = hour, y = sbi, group = sna, colour = sna)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) + theme(text = element_text(family = 'STHeiti')) 

dat %>% group_by(hour, sna) %>%
  summarise(mean(sbi)) %>%
  ggplot(aes(x = hour, y = `mean(sbi)`, group = sna, colour = sna)) +
  geom_line() + theme(text = element_text(family = 'STHeiti')) 

## 檢定四車站(sna)的平均租借數(sbi)是否相等
dat %>%
  ggplot(aes(x = sna, y = sbi)) +
  geom_boxplot() + theme(text = element_text(family = 'STHeiti')) 

## Ho: u1 = u2 = u3 = u4, H1: 全不相等
## results
fit <- aov(sbi ~ sna, data = dat) 
fit

## summary table: 
# WARNING: R provides Type I sequential SS, not the default Type III marginal SS reported by SAS and SPSS. 
# Rej Ho, 
summary(fit)

par(mfrow=c(2,2))
plot(fit)

# post-hoc test: Multiple Comparisons
TukeyHSD(fit)
```

### MANOVA

- Quiz: 同時對四車站(sna)的平均租借數(sbi)及空位數(bemp)進行檢定**

```{r}
dat <- ubike %>%
  filter(sna %in% c('世貿二館', '世貿三館', '信義廣場(台北101)', '市民廣場')) %>%
  select(date, hour, sna, sbi, bemp) 

head(dat)
summary(dat)

# manova
fit <- manova(cbind(sbi,bemp) ~ sna, data = dat) 

summary(fit) 

# 等同做兩次 anova
summary.aov(fit) # univariate ANOVA tables
summary(fit, test = "Wilks") # ANOVA table of Wilks' lambda
```

## Linear Model

### Simple Linear Model

- 簡單線性迴歸 (Linear Regression)
    - `y = ax + b`, `y = f(x) + e`
    - Assumptions:
        - Linear relationship.
        - Multivariate normality.
        - No or little multicollinearity.
        - No auto-correlation.
        - Homoscedasticity.
    - References
        - [Quick-R: Multiple (Linear) Regression](https://www.statmethods.net/stats/regression.html)
- Issues:
    - Data Transformation
        - `log(x)`, `x^2`
        - [Why do we log variables in regression model?](https://www.quora.com/Why-do-we-log-variables-in-regression-model)
            - skewness
            - transform a non-linear model into a linear model
            - economics interpretation
    - Missing Data
        - `na.omit`
        - replace with `mean()`
        - [Imputing Missing Data with R; MICE package](https://datascienceplus.com/imputing-missing-data-with-r-mice-package/)
    - Variable Selection
        - stepwise selection (forward, backward, both) 
    - Cross Validation
        - Bootstraping
    
- Quiz: 試探討各站平均租借數(sbi)影響因素

```{r}
dat <- ubike %>% filter(sna == '世貿二館') 

head(dat)
summary(dat)

## correlation
dat %>% select(sbi, temp, humidity, pressure) %>% cor() %>% knitr::kable()

# https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
# install.packages('corrplot')
library(corrplot)
dat %>% select(sbi, temp, humidity, pressure) %>% cor() %>% corrplot()

# Model #####
fit <- lm(sbi ~ humidity, data = dat)
summary(fit)

# useful functions for the fitted model #####
influence(fit) %>% str()  # influence values
coefficients(fit) # model coefficients
confint(fit, level=0.95) # CIs for model parameters 

# fitted values & residuals
fitted(fit)[1:10] # predicted values
residuals(fit)[1:10] # residuals

## diagnostic plots  #####
par(mfrow=c(2,2))
plot(fit)
```


- 迴歸診斷 (Regression Diagnostics)
    - [REGRESSION DIAGNOSTICS - John Fox](https://socialsciences.mcmaster.ca/jfox/Courses/Brazil-2009/index.html)
        - Data craft: examining and **transforming** variables.
        - Unusual data in linear models: **outliers**, leverage points, and influential observations, and what to do about them.
        - **Non-normality**, non-constant error variance, and nonlinearity in linear models: methods of detection, transformation, and other strategies.
        - Diagnostics for unusual data and nonlinearity in generalized linear models.
    - References
        - [Understanding Diagnostic Plots for Linear Regression Analysis](https://data.library.virginia.edu/diagnostic-plots/)
        - [Regression diagnostic plots - Prof. Chouldechova](https://www.andrew.cmu.edu/user/achoulde/94842/homework/regression_diagnostics.html#the-ideal-case-2)
        - [Quick-R: Regression Diagnostics](https://www.statmethods.net/stats/rdiagnostics.html)
        - [統計模型與迴歸分析 - 吳漢銘](http://www.hmwu.idv.tw/web/R/B05-hmwu_R-Regression.pdf)

```{r}
## compare results with aov function  #####
anova(fit) # anova table 
aov(sbi ~ humidity, data = dat) %>% summary()

## discussions #####
lm(temp ~ pressure, data = dat) %>% summary()
lm(temp ~ pressure + humidity, data = dat) %>% summary()
lm(sbi ~ temp + pressure + humidity, data = dat) %>% summary()
```

## Generalized Linear Models 

- 廣義線性模型
    - 是傳統線性模型的延伸，可使用非常態分布的資料(例如視為離散的計數或測量比例) 建立模型。 
    - `?glm`
        - http://www.theanalysisfactor.com/r-tutorial-glm1/
        - http://stats.idre.ucla.edu/r/
        - http://www.r-tutor.com/elementary-statistics/logistic-regression/estimated-logistic-regression-equation

- 羅吉斯迴歸 (Logistic regression)
    - 主要用於依變數 (y) 為二類 (0, 1) 的類別變項之情形 (即 y 符合二項分配 binomial distribution)
    - 實務常用於建立**是否選擇**之模型
    - Generalized Linear Models using binomial family

- Quiz: 試探討市民到站借車時，有車機率之影響因素

```{r}
### building analyse dataset 

# determine threshold value using histograme
ubike %>% select(sarea, sna, tot, bemp) %>%
  mutate(hv_cars = round(bemp/tot, digits = 4)) %>%
  ggplot(aes(x = hv_cars)) + 
  geom_histogram(binwidth = 0.05) 

# hv_cars rate >60%，hv_cars_yn is 1 (TRUE)
dat <- ubike %>% 
       mutate(hv_cars = round(bemp/tot, digits = 4),
              hv_cars_yn = ifelse(hv_cars > 0.6, 1, 0))

dat <- dat %>% filter(sarea == '信義區')

head(dat)

summary(dat)

### Cross-Analysis
reshape2::dcast(formula = sna ~ hv_cars_yn, dat)

reshape2::dcast(formula = sna ~ hv_cars_yn, dat, 
      fun.aggregate = function(x) round(mean(x), digit = 4), 
      value.var = 'hv_cars')

### Modeling: Logistic Regression
logitreg <- glm(hv_cars_yn ~ temp + humidity + pressure, data = dat, family = "binomial")
summary(logitreg)

## CIs using profiled log-likelihood
confint(logitreg)

## CIs using standard errors
confint.default(logitreg)

## odds ratios 
exp(coef(logitreg))

## Likelihood
logLik(logitreg)
```
